---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# splice2neo

<!-- badges: start -->
[![R-CMD-check](https://github.com/TRON-Bioinformatics/splice2neo/workflows/R-CMD-check/badge.svg)](https://github.com/TRON-Bioinformatics/splice2neo/actions)
[![Codecov test coverage](https://codecov.io/gh/TRON-Bioinformatics/splice2neo/branch/master/graph/badge.svg)](https://codecov.io/gh/TRON-Bioinformatics/splice2neo?branch=master)
`r badger::badge_devel("TRON-Bioinformatics/splice2neo", "blue")`
`r badger::badge_lifecycle("experimental", "blue")`
`r badger::badge_last_commit("TRON-Bioinformatics/splice2neo")`
<!-- badges: end -->


## Overview

This package provides functions for the analysis of splice
junctions and their association with somatic mutations. It integrates the output
of several tools which predict splicing effects from mutations or which detect
expressed splice junctions from RNA-seq data into a standardized splice junction
format based on genomic coordinates. Detected splice junctions can be filtered
against canonical splice junction and annotated with affected transcript sequences, CDS,
and resulting peptide sequences. Splice2neo currently supports splice events 
from alternative 3'/5' splice sites, exons skipping, intron retentions, 
exitrons and mutually exclusive exons.  
Integrating splice2neo functions and detection rules based on splice effect scores 
and RNA-seq support facilitates the identification of mutation-associated splice 
junctions which are tumor-specific and can encode neoantigen candidates. 

Documentation: https://tron-bioinformatics.github.io/splice2neo/

For a more detailed description and a full example workflow see the [vignette](https://tron-bioinformatics.github.io/splice2neo/articles/splice2neo_workflow.html)
 
## Installation

The R package splice2neo is not yet on [CRAN](https://CRAN.R-project.org) or
[Bioconductor](https://www.bioconductor.org/). Therefore, you need to install splice2neo
from [github](https://github.com/TRON-Bioinformatics/splice2neo). 

```{r, eval=FALSE}
if (!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes")
}

remotes::install_github("TRON-Bioinformatics/splice2neo")
```

## Example usage


This is a basic example of how to use some functions.

```{r}
library(splice2neo)

# load human genome reference sequence
requireNamespace("BSgenome.Hsapiens.UCSC.hg19", quietly = TRUE)
bsg <- BSgenome.Hsapiens.UCSC.hg19::BSgenome.Hsapiens.UCSC.hg19

```

### 1 Example data

We start with some example splice junctions provided with the package.
 
```{r}
junc_df <- dplyr::tibble(
  junc_id = toy_junc_id[c(1, 6, 10)]
)

junc_df
```

### 2 Add transcripts 

Next, we find the transcripts which are in the same genomic region as the splice junction and that may be affected by the junction.

```{r}
junc_df %>% 
  add_tx(toy_transcripts) %>% 
  head()

```


### 3 Modify transcripts with junctions

We modify the canonical transcripts by introducing the splice junctions. Then we
add the transcript sequence in a fixed-sized window around the junction
positions, the context sequence.


```{r}
toy_junc_df %>% 
  head()


toy_junc_df %>% 
  add_context_seq(transcripts = toy_transcripts, size = 400, bsg = bsg) %>% 
  head()

```

### 4 Annotate peptide sequence

Here, we use the splice junctions to modify the coding sequences (CDS) of the
reference transcripts. The resulting CDS sequences are translated into protein
sequence and further annotated with the peptide around the junction, the
relative position of the splice junction in the peptide, and the location of the
junction in an open reading frame (ORF).

```{r}


toy_junc_df %>% 
  
  # add peptide sequence
  add_peptide(cds=toy_cds, flanking_size = 13, bsg = bsg) %>% 

  # select subset of columns
  dplyr::select(junc_id, peptide_context, junc_in_orf, cds_description) %>% 
  head()

```


## Issues

Please report issues here: https://github.com/TRON-Bioinformatics/splice2neo/issues

## Citation

Lang, Franziska, Patrick Sorn, Martin Suchan, Alina Henrich, Christian Albrecht, Nina Koehl, Aline Beicht, et al. 2023. “Prediction of Tumor-Specific Splicing from Somatic Mutations as a Source of Neoantigen Candidates.” bioRxiv. https://doi.org/10.1101/2023.06.27.546494.
